Hydra Battles and Cie

### 2021-08-05 doc/part-hydras.tex
converting chapter 3 to alectryon (continued)

\begin{Coqsrc}
(** multiplication of an ordinal by a natural number *)

Definition mult_fin_r  (alpha : t) (p : nat): t :=
  match alpha, p with
 |  (0,0), _  => zero
 |  _, 0 => zero
 |  (0, n), p => (0, n * p)
 |  ( n, b),  n' => ( n *  n', b)
 end.
Infix "*" := mult_fin_r : o2_scope.

(** multiplication of  a natural number by an ordinal *)

Definition mult_fin_l (n:nat)(alpha : t) : t :=
  match n, alpha with
 |  0, _  => zero
 |  _, (0,0) => zero
 |   n , (0,n') => (0, (n*n')%nat)
 |  n, (n',p') => (n', (n * p')%nat)
 end.

Example e1 : (omega * 7 + 15) * 3 = omega * 21 + 15.
Proof. reflexivity. Qed.

Example e2 :  mult_fin_l 3 (omega * 7 + 15) = omega * 7 + 45.
Proof. reflexivity. Qed.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma unique_decomposition alpha : 
    exists! i j: nat,  alpha = omega * i + j.
\end{Coqsrc}

@@

\begin{Coqsrc}
Section Merge.

  Variable A: Type.

  Local Definition m (p : list A * list A) :=
    omega * length (fst p) + length (snd p).

  Function  merge  (ltb: A -> A -> bool)
          (xys: list A * list A)
          {wf (measure_lt m) xys} :
    list A :=
    match xys with
      (nil, ys) => ys
    | (xs, nil) => xs
    | (x :: xs, y :: ys) =>
      if ltb x y then x :: merge  ltb (xs, (y :: ys))
      else y :: merge  ltb ((x :: xs), ys)
    end.

  - intros; unfold m, measure_lt; cbn; destruct xs0; simpl; left; lia.
  - intros; unfold m, measure_lt; cbn; destruct ys0; simpl; right; lia.
  - apply wf_measure.
  Defined.

End Merge.

Goal forall l, merge nat Nat.leb (nil, l) = l.
  intro; now rewrite merge_equation.
Qed.
\end{Coqsrc}

@@

 Variable m : Hydra -> ON_Omega2.t.

 Context (Hvar : Hvariant (ON_Generic.wf (ON:=Omega2))  free m).
\end{Coqsrc}

@@

\begin{Coqsrc}
 Let iota (p: ON_Omega2.t) := 
     node (hcons_mult (hyd1 head) (fst p)
                    (hcons_mult head (snd p) hnil)).
\end{Coqsrc}

@@

\begin{Coqsrc}
   Let big_h := hyd1 (hyd2 head head).  
\end{Coqsrc}

@@

\begin{Coqsrc}
   Let small_h := iota (m big_h).
\end{Coqsrc}

@@

\begin{Coqsrc}
  Lemma big_to_small: big_h -+-> small_h.
\end{Coqsrc}

@@

\begin{Coqsrc}
   Corollary m_lt : m small_h o< m big_h.
\end{Coqsrc}

@@

\begin{itemize}
\item The proof of the lemma \texttt{m\_ge : m big\_h  o<= m small\_h} uses well-founded induction on \texttt{big\_h}.

\item For any pair $p$, we have to distinguish between three cases, according to the value of $p$'s components.
  \begin{itemize}
  \item $p=(0,0)$
  \item $p=(i,0)$, where $i>0$\,: $p$ corresponds to a limit ordinal
  \item $p=(i,j)$, where $j>0$\,: $p$ is the successor of $(i,j-1)$.
  \end{itemize}
\end{itemize}

Let us define the notion of elementary ``step'' of decreasing sequences in
\texttt{t}

\begin{Coqsrc}
  Inductive step : t -> t -> Prop :=
  | succ_step : forall i j,  step (i, S j) (i, j)
  | limit_step : forall i j, step (S i, 0) (i, j).
\end{Coqsrc}

The following lemma establishes a correspondance between the relation
\texttt{step} and hydra battles.

\begin{Coqsrc}
  Lemma step_to_battle : forall p q, step p q -> iota p -+-> iota q.
\end{Coqsrc}

Thus, starting from any inequality $q < p$ on type \texttt{t}, we can build 
by \emph{transfinite induction} (\emph{i.e.} well-founded) over \texttt{p} a battle 
that transforms the hydra $\iota(p)$ into $\iota(q)$.

\vspace{4pt}
\emph{From Module~\href{../theories/html/hydras.Hydra.Omega2_Small.html\#m_ge}{Hydra.Omega2\_Small}}

\begin{Coqsrc}
  Lemma m_ge : forall p : t,   p o<= m (iota p).
  Proof.
    unfold small_h; pattern (m big_h) .   
     apply  well_founded_induction with (R := lt) (1:= lt_wf).
     intro p ; pattern p;
     apply  well_founded_induction with 
               (R := lt2) (1:= wf_lexico lt_wf lt_wf);
     intros (i,j) IHij. 
\end{Coqsrc}

\begin{Coqanswer}
  i, j : nat
  IHij : forall y : t, y o< (i, j) -> y o<= m (iota y)
  ============================
  (i, j) o<= m (iota (i, j)) 
\end{Coqanswer}

Then we have  three cases to consider, according to the value of $p$.
\begin{itemize}
\item If $p=(0,0)$ then obviously, $\iota(p)\geq p = (0,0)$
\item If  $p=(i+1,0)$ for some $i\in\mathbb{N}$, we
 remark  that $p$ is strictly greater than any pair $ (i, j)$, where $j$ 
is any natural number.

Applying the battle rules, for any $j$, we have $\iota(i+1,j)  {\round} \iota(i, j) $, thus $m(\iota(p)) > m(\iota(i,j)$ since  $m$ is assumed to be a variant.

Applying the induction hypothesis, we get the inequality
 $ m(\iota(i,j)) \geq (i,j)$ for any $j$.

Thus, $m(\iota(p)) > (i,j)$ for any $j$.
Applying the lemma \texttt{limit\_is\_lub}, we get  the inequality
$m(\iota(i+1,0))\geq (i+1,0)$

\item If $p=(i,j+1)$ with $j\in\mathbb{N}$, we have  $\iota(p)  {\round} \iota(i, j) $,
hence $m(\iota(p))> m(\iota(i,j)) \geq (i,j)$, thus $m(\iota(p))\geq (i,j+1)=p$

\begin{Coqsrc}
  (* ... *)
Qed.
\end{Coqsrc}

@@

\vspace{4pt}
\noindent
In \coq (Module~\href{../theories/html/hydras.Hydra.Omega2_Small.html\#Impossible}{Hydra.Omega2\_Small}):

\begin{Coqsrc}
  Theorem Impossible: False.
  Proof.
    destruct (StrictOrder_Irreflexive (m big_h)).
    apply le2_lt2_trans with (m small_h).
    -  unfold small_h; apply m_ge.
    -  apply m_lt. 
Qed. 
End Impossibility_Proof.
\end{Coqsrc}

>>>>

\input{movies/snippets/ON_Omega2/multFinDef}

@@

\input{movies/snippets/ON_Omega2/uniqueDecomposition}

@@

\input{movies/snippets/ON_Omega2/Merge}

@@

\input{movies/snippets/Omega2_Small/Impossibility}
\input{movies/snippets/Omega2_Small/Impossibilitya}

@@

\input{movies/snippets/Omega2_Small/Impossibilityc}

@@

\input{movies/snippets/Omega2_Small/Impossibilityb}

@@

\input{movies/snippets/Omega2_Small/Impossibilityd}

@@

\input{movies/snippets/Omega2_Small/bigToSmall}

@@

\input{movies/snippets/Omega2_Small/mLt}

@@

The first steps of our proof prepare a well-founded indsction on $\omega^2$.

\input{movies/snippets/Omega2_Small/mGe}

\begin{Coqanswer}
  i, j : nat
  IHij : forall y : t, y o< (i, j) -> y o<= m (iota y)
  ============================
  (i, j) o<= m (iota (i, j))

Then a case analysis on $i$ and $j$ allows us to
consider three cases :

\begin{itemize}
\item $i=j=0$: the inequality is trivial.
\item $i=1+l, j=0$ ($(i,j)$ is a limit ordinal): By the induction hypothesis \texttt{IHij},
  $(l,k)\leq m(\iota(l,k))$ for any $k$. But (by the rules of the hydra game), $\iota(i,0)$ is transformed into any $\iota(l,k)$ in one round. Thus $m(\iota(l,k)) < m(\iota(i,0))$ for any $k$.
  Therefore, $(l,k) <  m(\iota(i,0))$ for any $k$, thus
  $(i,0) \leq m(\iota(i,0))$.
 \item $j= l+1$  ($(i,j)$ is a successor).  a similar, but simpler case: we apply the induction hypothesis to the pair $(i,l)$.
 \end{itemize}

 Please look at the proof script for more details.

 \input{movies/snippets/Omega2_Small/mGeb}

@@

\vspace{4pt}
\noindent
\emph{From Module~\href{../theories/html/hydras.Hydra.Omega2_Small.html\#Impossible}{Hydra.Omega2\_Small}}

  \input{movies/snippets/Omega2_Small/Impossible}
